Acceleration of plague outbreaks in the second pandemic\\\emph{Electronic \supp}

### 2020-07-11 supp.Rnw
a few minor tweaks / temporary comments

For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies \cite{KeelGill00}.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

@@

Keeling and Gilligan \cite[p.\,2226]{KeelGill00b} relate the basic
reproduction number to rat density via
%
\begin{linenomath*}
\begin{equation}\label{E:KGR0}
\R_0 = \frac{\beta_{\rm R} K_{\rm F}}{d_{\rm F}}\Big[
1 - \exp{(-aK_{\rm R})} \Big] \,,
\end{equation}
\end{linenomath*}
%
where $\beta$ and $K$ denote transmission rate and carrying capacity, respectively, and the subscripts F and R denote fleas and rats, respectively.  \cref{E:KGR0} implies
%
\begin{linenomath*}
\begin{equation}\label{E:R0ratio}
\frac{\R_{0,2}}{\R_{0,1}} = 
\frac{1 - \exp{(-aK_{{\rm R},2})}}{1 - \exp{(-aK_{{\rm R},1})}} \,,
\end{equation}
\end{linenomath*}
%
which, for $aK_{{\rm R},j}\ll1$, simplifies to
%
\begin{linenomath*}
\begin{equation}\label{E:aKsmall}
\frac{\R_{0,2}}{\R_{0,1}}
\;\approx\; 
\frac{K_{{\rm R},2}}{K_{{\rm R},1}}
\quad.
\end{equation}
\end{linenomath*}
%
The approximation (appropriate in the limit $aK\to0$) provides a lower
bound to the true relationship, indicating that rat density would have
had to increase by at least a factor of  to account for a similar
increase in $\R_0$.

\cref{F:KvsR} shows the exact relationship \eqref{E:R0ratio} for
several values of $aK_{{\rm R},1}$, together with the approximation
for small $aK$ \eqref{E:aKsmall}.  Keeling and Gilligan \cite[Table 1,
p.\,2221]{KeelGill00b} adopt values of $a$ and $K_{{\rm R}}$ that
yield $aK_{{\rm R}}=10$.  If this is the correct order of magnitude
for $aK_{{\rm R},1}$ then \cref{E:R0ratio} implies that no increase in
rat density would be sufficient to yield an increase in $\R_0$ by a
factor of \Sexpr{round(R0ratio,1)} (or any factor detectably greater than 1).

@@

As a vector-borne disease, the rate of bubonic plague spread is primarily affected by the flea-rat ratio.
In the traditional Ross-MacDonald model for vector-borne epidemics, $\rzero$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host). In the rat-flea model, $\rzero$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
From Keeling and Gilligan's plague model \cite{KeelGill00b}, the expected change in $r$ for a change in flea-rat ratio from $K_{F0}$ to $K_F$ is $(K_F/K_{F0})^S$, where $S$ is the sensitivity ($\approx 1.5$). Thus in order to a see a fivefold increase in $r$ we would need a change of $5^{(1/S)} \approx \Sexpr{round(5^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.

@@

\begin{table}[h!]
\caption{Observed cross-correlation and delay between London Bills of
  Mortality (LBoM) records and other sources. All lags measured in
  weeks.  Rows corresponding to wills data are highlighted in grey.}
\begin{center}
  \input{analysis/tables/wills_delay_tab.tex}
\end{center}
\label{tab:delays}
\end{table}

@@

Robustly fitting phenomenological epidemic curves to small, noisy data
sets proved to be surprisingly challenging\footnote{See ``monsters in
  the basement'',
  \url{https://redpenblackpen.tumblr.com/post/145304820562/monsters-in-the-basement}.}. In
the course of developing the full model we experimented with a range
of optimizers --- specifically the BFGS and Nelder-Mead options for
R's \code{optim} function as well as R's \code{nlminb} function; we
also tried an approach that iterated back and forth between
Nelder-Mead and BFGS until convergence was achieved or the fit was
sufficiently stable. We chose \code{nlminb} because it gave the best
results (highest log likelihood) for point estimates.
In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e. fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$. However, we found that the scaled-tangent link declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).

>>>>

For pneumonic plague, sufficiently detailed data exist for a number of
modern outbreaks to allow an estimate of the latent and infectious
periods (and hence of the generation interval distribution)
\cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given
value of $r$ \cite{WallLips07}. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies \cite{KeelGill00}.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

@@

Keeling and Gilligan \cite[p.\,2226]{KeelGill00b} relate the basic
reproduction number to rat density via
%
\begin{linenomath*}
\begin{equation}\label{E:KGR0}
\R_0 = \frac{\beta_{\rm R} K_{\rm F}}{d_{\rm F}}\Big[
1 - \exp{(-aK_{\rm R})} \Big] \,,
\end{equation}
\end{linenomath*}
%
where $\beta$ and $K$ denote transmission rate and carrying capacity, respectively, and the subscripts F and R denote fleas and rats, respectively.  \cref{E:KGR0} implies
%
\begin{linenomath*}
\begin{equation}\label{E:R0ratio}
\frac{\R_{0,2}}{\R_{0,1}} = 
\frac{1 - \exp{(-aK_{{\rm R},2})}}{1 - \exp{(-aK_{{\rm R},1})}} \,,
\end{equation}
\end{linenomath*}
%
which, for $aK_{{\rm R},j}\ll1$, simplifies to
%
\begin{linenomath*}
\begin{equation}\label{E:aKsmall}
\frac{\R_{0,2}}{\R_{0,1}}
\;\approx\; 
\frac{K_{{\rm R},2}}{K_{{\rm R},1}}
\quad.
\end{equation}
\end{linenomath*}
%
The approximation (appropriate in the limit $aK\to0$) provides a lower
bound to the true relationship, indicating that rat density would have
had to increase by at least a factor of \david{???missing number???}
to account for a similar increase in $\R_0$.

\cref{F:KvsR} shows the exact relationship \eqref{E:R0ratio} for
several values of $aK_{{\rm R},1}$, together with the approximation
for small $aK$ \eqref{E:aKsmall}.  Keeling and Gilligan \cite[Table 1,
p.\,2221]{KeelGill00b} adopt values of $a$ and $K_{{\rm R}}$ that
yield $aK_{{\rm R}}=10$.  If this is the correct order of magnitude
for $aK_{{\rm R},1}$ then \cref{E:R0ratio} implies that no increase in
rat density would be sufficient to yield an increase in $\R_0$ by a
factor of \Sexpr{round(R0ratio,2)} (or any factor detectably greater
than 1).

@@

As a vector-borne disease, the rate of bubonic plague spread is primarily affected by the flea-rat ratio.
In the traditional Ross-MacDonald model for vector-borne epidemics, $\rzero$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host). In the rat-flea model, $\rzero$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
From Keeling and Gilligan's plague model \cite{KeelGill00b}, the
expected change in $r$ for a change in flea-rat ratio from $K_{\rm
  F0}$ to $K_{\rm F}$ is $(K_{\rm F}/K_{\rm F0})^S$, where $S$ is the
sensitivity ($\approx 1.5$). Thus in order to a see a fivefold
\david{why FIVEfold?  we need only FOURfold; didn't we have a macro
  for this?} increase in $r$ we would need a change of $5^{(1/S)} \approx \Sexpr{round(5^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.

@@

\begin{table}[h!]
\caption{Observed cross-correlation and delay between London Bills of
  Mortality (LBoM) records and other sources. All lags measured in
  weeks.  Rows corresponding to wills data are highlighted in grey.
  \david{why is the row shading wrong?}
}
\begin{center}
  \input{analysis/tables/wills_delay_tab.tex}
\end{center}
\label{tab:delays}
\end{table}

@@

Robustly fitting phenomenological epidemic curves to small, noisy data
sets proved to be surprisingly challenging\footnote{See ``monsters in
  the basement'',
  \url{https://redpenblackpen.tumblr.com/post/145304820562/monsters-in-the-basement}.}. In
the course of developing the full model we experimented with a range
of optimizers --- specifically the BFGS and Nelder-Mead options for
R's \code{optim} function as well as R's \code{nlminb} function; we
also tried an approach that iterated back and forth between
Nelder-Mead and BFGS until convergence was achieved or the fit was
sufficiently stable. We chose \code{nlminb} because it gave the best
results (highest log likelihood) for point estimates.
In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e., fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$. However, we found that the scaled-tangent link declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).
